---
title: "Basic workflow to work with ERA5land data with stars"
author: "Carles Milà"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', message = F, warnings = F)
```

<style type="text/css">
pre {
  font-size: 11px
}
</style>

## Getting ready

This notebook contains a basic example of a R-based workflow for data management
and analysis of ERA5land data with the stars package. The example data consist on
2m and skin temperature ERA5land data for the study area comprised between 
40ºN, 44ºN, 0ºE, 4ºE at an hourly resolution for January 2020. These can be easily
downloaded from the [climate data store](https://cds.climate.copernicus.eu/#!/home) and 
are also included in the repo.

```{r libraries}
library("stars")
library("sf")
library("readr")
library("dplyr")
library("viridis")
library("ggplot2")
```

## Read data and stars 101

We easily read ERA5land data using `read_stars`.

```{r read ERA5}
(airtemp <- read_stars("data/ERA5Land_2mtemp.grib"))
(skintemp <- read_stars("data/ERA5Land_skintemp.grib"))
```

We use read our AOI that was stored as a rds object.

```{r read AOI, fig.height=3}
(catalonia <- read_rds("data/cat.rds"))
plot(catalonia)
```

We can merge the two products and fix their names.

```{r fix names}
alltemp <- c(airtemp, skintemp)
names(alltemp) <- gsub(".grib", "", names(alltemp))
alltemp
```

We subset stars objects by using []. 1st is the attribute, next are the dimensions in order.

```{r subset}
alltemp[1] # subset attribute
alltemp[, 1] # subset 1st dimension
alltemp[,, 1] # subset 2nd dimension
alltemp[,,, 1] # subset 3rd dimension
```

We now fix the time dimension with the correct times we downloaded. For these data, it's hourly for Jan 2020.

```{r fix time}
datetime_era5 <- seq.POSIXt(ISOdatetime(2020, 1, 1, 0, 0, 0),
                            ISOdatetime(2020, 1, 31, 23, 0, 0), by="1 hour")
dates_era5 <- seq.Date(as.Date("2020-01-01"), as.Date("2020-01-31"), "1 day")
(alltemp <- st_set_dimensions(alltemp, 3, names="datetime", values=datetime_era5))
```

We can use the basic plotting capabilities of `stars` to check the data.

```{r}
plot(alltemp["ERA5Land_2mtemp",,,1:10], col=viridis(5))
plot(alltemp["ERA5Land_skintemp",,,1:10], col=viridis(5))
```


## Band calculations 

We can now make easy calculations within/between bands now. Here we do a ratio of air/skin temp.

```{r band calcs}
(alltemp <- mutate(alltemp, ratio = ERA5Land_2mtemp/ERA5Land_skintemp))
plot(alltemp[3,,,1:10], col=viridis(5))
```


## Time aggregation and filtering

Now we can easily calculate daily averages, we convert the time dimension to date and aggregate.

```{r time aggregation}
(alltemp <- aggregate(alltemp, "day", FUN=mean))
(alltemp <- st_set_dimensions(alltemp, "time", names = "date", values = dates_era5))
plot(alltemp[,1:10,,], col=viridis(5))
```

We can also filter by dimensions, for example date, if we want to:

```{r}
(alltemp_8days<- dplyr::filter(alltemp, date <= as.Date("2020-01-08")))
```

## Transforming CRS and cropping to AOI

Now we transform from the geographic to our target CRS. To do that, we want to use bilinear interpolation using GDAL. To do that, we need to create a template first and then transform using GDAL one attribute at a time.

```{r}
alltemp_template <- st_warp(alltemp, crs = st_crs(catalonia)) # This uses nearest neighbour
# When using GDAL, we can only transform one attribute at a time
alltemp <- c(st_warp(alltemp["ERA5Land_2mtemp"], alltemp_template, method = "bilinear", use_gdal = T),
             st_warp(alltemp["ERA5Land_skintemp"], alltemp_template, method = "bilinear", use_gdal = T),
             st_warp(alltemp["ratio"], alltemp_template, method = "bilinear", use_gdal = T))
# When using GDAL, names and time dimension info are lost, so we redefine them
names(alltemp) <- names(alltemp_template)
alltemp <- st_set_dimensions(alltemp, 3, names = "date", values = dates_era5)
```

Now we can crop to our study area.

```{r}
alltemp <- st_crop(alltemp, catalonia)
```

## Using ggplot2

We can use stars objects in ggplot2 using `geom_stars` and use time dimension
for faceting. Using `tmap` is also possible.

```{r}
ggplot() +
  geom_stars(data=alltemp["ratio",,,1:8]) +
  facet_wrap(~ date, nrow = 2) +
  scale_fill_continuous(type = "viridis", na.value = "#FFFFFF00") +
  xlab("") + ylab("") +
  theme_bw() +
  theme(aspect.ratio=1, axis.ticks = element_blank(), axis.text = element_blank()) 
```

## Extract pixel values

To extract pixel values at certain points, the raster stars objects is converted to a stars vector object, which
we can easily transform into a data frame.

```{r}
catalonia_centroid <- st_centroid(catalonia)
(alltemp_pxls <- st_extract(alltemp, catalonia_centroid))
alltemp_pxls <- as.data.frame(alltemp_pxls)
head(alltemp_pxls)
```

## Zonal statistics

If we want to calculate statistics by area, we'll use the `aggregate` function. Here we compute the mean value in our study area.

```{r}
(zonal_means <- aggregate(alltemp, catalonia, FUN=mean, na.rm=T))
zonal_means <- as.data.frame(zonal_means)
head(zonal_means)
```
